\ProjectName{}: Data Imputation as a Query Optimization

### 2017-02-23 paper/introduction.tex
paper: some typos, clarifications, fixes
paper: work on intro

Handling incorrect or dirty data is a complex and challenging problem for data scientists.
One way in which a dataset can be dirty is for parts of it to be missing altogether.
Missing data is one of the simpler variants of dirty data, but if handled naively, can still cause analyses to be incorrect.
To handle this problem, users may manually clean their dataset by performing
some statistical analysis to replace missing data elements with likely values.
This process is called \emph{imputation}.
Traditional imputation methods replace all missing values in a dataset to create a new clean dataset.
Although manual imputation solves the problem of missing data, in the age of big data it may be very expensive to run an imputation algorithm on an entire dataset.
Additionally, it may not even be necessary to completely clean the data to make it usable.
Some users may be willing to run queries on dirty data, simply ignoring the missing values, as long as they do not have to pay the cost of imputation.
Others may want to run queries on a subset of the data, so do not need to impute
every field in every record. Yet others may want to customize the 
imputation algorithm for the tradeoffs and demands of a particular domain.

In this paper, we present \ProjectName{}, a database system which is designed to interact with a dirty dataset as though it were clean.
The guiding design principle behind \ProjectName{} is that the user should never see missing data or have to modify their queries to account for it.
To achieve this goal, we perform imputation on the fly, during query execution.
Performing imputation at query time allows our system to impute only the data necessary to run the query, and it allows users to flexibly trade accuracy for computation time.

>>>>

% Context
Handling incorrect or dirty data is a complex and challenging problem for analysts.
% Problem
One way in which a dataset can be dirty is for parts of it to be missing altogether.
Missing data is one of the simpler variants of dirty data, but if handled naively, can still cause analyses to be incorrect.
According to some estimates, data scientists spend between 50 and 80 percent of their time performing tasks such as collecting and cleaning datasets~\cite{data-science-cleaning}.

% Solution
To handle this problem, users may manually clean their dataset by performing some statistical analysis to replace missing data elements with likely values.
This process is called \emph{imputation}.
We have developed a system called \ProjectName{} with a novel query optimization algorithm which allows imputation to occur on-the-fly, during query execution.
The guiding design principle behind \ProjectName{} is that the user should never see missing data or have to modify their queries to account for it.

% Prior work
There is relevant prior work in three broad areas: statistics, databases, and time-series forecasting.
Prior work in statistics has focused on imputation quality~\cite{burgette2010multiple} and runtime cost~\cite{akande2015empirical}.
Prior work in databases has a established simple semantics for handling tuples with missing values~\cite{codd1973understanding,grant1977null}.
Prior work in time-series forecasting has incorporated forecasting extensions into domain-specific databases and characterized their behavior~\cite{parisi2011embedding,parisi2013temporal,duan2007processing}.
In contrast, \ProjectName{} does not focus on the imputation algorithm itself but rather on the planning involved in its use in non-trivial query plans.

% Implications
Our approach enables an exploratory analysis workflow in which the analyst can issue standard SQL queries over a data set and get good results, even if that data has missing values.

% Background
Traditional imputation methods work over the entire data set and replace all missing values.
Although this solves the problem of missing data, running a sophisticated imputation algorithm over a large data set can be very expensive.
A simpler approach might drop all rows with missing values, but not only can this introduce bias into the results, but this may result in discarding all of the data.
Finally, for many queries, it is not necessary to clean the entire data set, and a significant speed up can be obtained by cleaning only the necessary parts.

% Scope
\ProjectName{} is designed to enable data exploration, by allowing analysts to run their queries and get good results without dealing with imputation.
However, we optimize query plans with respect to a heuristic measure of imputation quality, and provide no hard guarantees about the accuracy of the results.

\subsection{Contributions}
This paper makes the following contributions:
\begin{itemize}
\item \textbf{Relational algebra with imputation:}
  We extend the standard relational algebra with two new operators to represent imputation operations: impute ($\mu$) and drop ($\delta$).
  Impute fills in missing data values using a regression model (CE-CART)~\cite{burgette2010multiple}.
  Drop simply drops tuples which contain null values.
\item \textbf{Model of imputation quality and cost:}
  We extend the traditional cost model for query plans to incorporate a measure of the quality of the imputations performed.
  We use the cost model to abstract over the imputation algorithm used.
  To add an additional imputation technique, it is sufficient to characterize it with two functions: one to describe its running time and one to describe the quality of its results.
\item \textbf{Query planning with imputation:}
  We present a novel query planning algorithm which jointly optimizes for running time and the quality of the results.
  It does so by maintaining the set of Pareto optimal plans according to the cost model.
  By deferring selection of the final plan, we make it possible for the user to trade off running time and result quality.
\end{itemize}

% Summary of related work

% Although this manual imputation solves the problem of missing data, in the age of big data it may be very expensive to run an imputation algorithm on an entire dataset.
% Additionally, it may not be necessary to completely clean the data to make it usable.
% Some users may be willing to run queries on dirty data, simply ignoring any missing values, as long as they do not have to pay the cost of imputation.
% Others may want to run queries on a subset of the data, and so do not need to impute
% every field in every record. Yet others may want to customize the 
% imputation algorithm for the tradeoffs and demands of a particular domain.

% In this paper, we present \ProjectName{}, a database system which is designed to interact with a dirty dataset as though it were clean.
% To achieve this goal, we perform imputation on the fly, during query execution.
% Performing imputation at query time allows our system to impute only the data necessary to run the query, and it allows users to flexibly trade imputation quality for computation time.

### 2017-02-23 paper/relatedwork.tex
paper: work on intro

Imputation of missing values is a widely studied field within the statistics and machine
learning communities. As highlighted in \cite{gelman2006data}, missing data
can appear for a variety of reasons, including both random and conditioned on
existing values (observed and missing). Methods in the statistical community
focus on correctly modeling relationships between the attributes to factor in
varied forms of missingness. For example, Burgette and Reiter
(\cite{burgette2010multiple}) discuss the usage of sequential regression trees
for imputing missing data.

In \cite{akande2015empirical}, Akande et al analyze the performance of various
multiple imputation techniques on the American Community Survey dataset. 
The computational difficulties of imputing on large base
tables are well known and can limit approaches. For example, Akande finds that
one approach (MI-GLM) is prohibitively expensive when attempting to impute on data that
includes variables with potentially large domains (ten categories in their case).
In contrast, \ProjectName{} allows users to specify a tradeoff between
information lost and runtime performance, allowing users to perform queries
directly on the entire dataset. Furthermore,
the query planner's imputation is guided by the requirements of each specific
query's operators, rather than requiring broad assumptions about query
workloads.

\subsection{Missing Values and Databases}
There is a long history in the database community surrounding the
treatment of nulls. As early as 1973, \cite{codd1973understanding}
provides a treatment of the semantics of null. Multiple
papers have described various (at times conflicting) treatments
of nulls \cite{grant1977null}. \ProjectName's main design invariant - no relational operator
sees missing values for attributes it must operate on and users should never see
missing data - eliminates
the need to handle null value semantics, while guaranteeing soundness (modulo
imputation strategy) of the query evaluation.

>>>>

Imputation of missing values is a widely studied field within the statistics and machine
learning communities. As highlighted in~\cite{gelman2006data}, missing data
can appear for a variety of reasons, including both random and conditioned on
existing values (observed and missing). Methods in the statistical community
focus on correctly modeling relationships between the attributes to factor in
varied forms of missingness. For example, Burgette and Reiter~\cite{burgette2010multiple} discuss the usage of sequential regression trees
for imputing missing data.

In~\cite{akande2015empirical}, Akande et al analyze the performance of various
multiple imputation techniques on the American Community Survey dataset. 
The computational difficulties of imputing on large base
tables are well known and can limit approaches. For example, Akande finds that
one approach (MI-GLM) is prohibitively expensive when attempting to impute on data that
includes variables with potentially large domains (ten categories in their case).
In contrast, \ProjectName{} allows users to specify a tradeoff between
information lost and runtime performance, allowing users to perform queries
directly on the entire dataset. Furthermore,
the query planner's imputation is guided by the requirements of each specific
query's operators, rather than requiring broad assumptions about query
workloads.

\subsection{Missing Values and Databases}
There is a long history in the database community surrounding the
treatment of nulls. As early as 1973,~\cite{codd1973understanding}
provides a treatment of the semantics of null. Multiple
papers have described various (at times conflicting) treatments
of nulls~\cite{grant1977null}. \ProjectName's main design invariant - no relational operator
sees missing values for attributes it must operate on and users should never see
missing data - eliminates
the need to handle null value semantics, while guaranteeing soundness (modulo
imputation strategy) of the query evaluation.
